---
title: "lab_06"
author: "derek willis"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and any other libraries you'll need
library(tidyverse)
library(lubridate)

```

## Get Our PDF

We'll be working with the [Maryland monthly voter registration update](https://elections.maryland.gov/pdf/vrar/2024_09.pdf) from the State Board of Elections. You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to write a couple of paragraphs that summarize the changes in voter registration in Maryland during the month of September.

## Setup Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds.

This PDF has multiple possible tables to extract. We're going to make four dataframes: 

* new registrations by party
* removals by party
* changes in registration
* current registrations by county.

You will need to draw boxes around four sections and export four CSV files that you will load into R. In Tabula, draw a box around the borders of those tables and click the "Preview & Export Extracted Data" button for each one. The export should look pretty clean, but might be missing some column headers. You'll need to include total columns.

Export each CSV (each one should be called `tabula-2024-09.csv` by default, so you should rename them for clarity) to your lab_06/data folder.

From there, you will need to read in and clean up the data, and in particular the headers. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. The pre_lab has an example of how to do this!

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers must not contain spaces and they should have meaningful names, not "x1" or something similar. The current registrations dataframe *MUST* include the county name, but should not include the columns under `Changes` and `Party Affiliation From`. How you do that is up to you (and could involve something outside R), but you can use `select()` with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns.

```{r}
# Party registrations
party_registrations <- read_csv("party_affiliation_noheader.csv" , col_names = FALSE)
party_registrations <- party_registrations |>
  rename(COUNTY = X1)

party_registrations <- party_registrations |>
select(-X2, -X3)

party_registrations <- party_registrations |>
  rename(DEM = X4)

party_registrations <- party_registrations |>
  rename(REP = X5)

party_registrations <- party_registrations |>
  rename(NLM = X6)

party_registrations <- party_registrations |>
  rename(LIB = X7)

party_registrations <- party_registrations |>
  rename(CRN = X8)

party_registrations <- party_registrations |>
  rename(UNAF = X9)

party_registrations <- party_registrations |>
  rename(OTH = X10)

party_registrations <- party_registrations |>
  rename(TOTAL = X11)

```

```{r}
# Removals
removals <- read_csv("removals.csv" , col_names = FALSE)
removals <- removals |>
  select(-X1, -X2, -X3, -X4, -X5, -X6, -X7, -X8, -X9, -X10, -X11,)
removals <- removals |>
rename(SUMMARY = X12)
removals <- removals |>
rename(DEM = X13)
removals <- removals |>
rename(REP = X14)
removals <- removals |>
rename(NLM = X15)
removals <- removals |>
rename(LIB = X16)
removals <- removals |>
rename(GRN = X17)
removals <- removals |>
rename(UNAF = X18)
removals <- removals |>
rename(OTH = X19)
removals <- removals |>
rename(TOTAL = X20)
removals <- removals |>
  filter(!is.na(DEM))

```

```{r}
# Changes
changes <- read_csv("changes.csv" , col_names = FALSE)
changes <- changes |>
  rename(COUNTY = X1)
changes <- changes |>
  rename(ADDRESS = X2)
changes <- changes |>
  rename(NAME = X3)
```

```{r}
# Current registrations
current <- read_csv("current.csv" , col_names = FALSE)
current <- current |>
  select(-X2, -X3, -X4, -X5, -X6, -X7, -X8, -X9, -X10, -X11, -X20,-X21)

current <- current |>
  rename(COUNTY = X1)
  current <- current |>
  rename(DEM = X12)
  current <- current |>
  rename(REP = X13)
  current <- current |>
  rename(NLM = X14)
  current <- current |>
  rename(LIB = X15)
  current <- current |>
  rename(GRN = X16)
  current <- current |>
  rename(UNAF = X17)
  current <- current |>
  rename(OTH = X18)
  current <- current |>
  rename(TOTAL = X19)
  
```
## Rinse, Repeat: September 2022 data

Do all of that again, for the [September 2022 voter registration update](https://elections.maryland.gov/pdf/vrar/2022_09.pdf), so we can compare the two. If you've done well, this is easy: just do the same steps in Tabula, export the files and use the same code to read in and clean up the data, *just remember to give these dataframes different names* so that you can compare September 2024 to September 2022.

To answer the following questions, you'll need to calculate percentages and do other calculations. Whether you do that by combining or joining dataframes or you write code to run separately on both is up to you.

```{r}
# Party registrations
party_registrations_2022 <- read_csv("party_affiliation_22.csv" , col_names = FALSE)
party_registrations_2022 <- party_registrations_2022 |>
  rename(COUNTY = X1)

party_registrations_2022 <- party_registrations_2022 |>
select(-X2, -X3)

party_registrations_2022 <- party_registrations_2022 |>
  rename(DEM = X4)

party_registrations_2022 <- party_registrations_2022 |>
  rename(REP = X5)

party_registrations_2022 <- party_registrations_2022 |>
  rename(NLM = X6)

party_registrations_2022 <- party_registrations_2022 |>
  rename(LIB = X7)

party_registrations_2022 <- party_registrations_2022 |>
  rename(CRN = X8)

party_registrations_2022 <- party_registrations_2022 |>
  rename(UNAF = X9)

party_registrations_2022 <- party_registrations_2022 |>
  rename(OTH = X10)

party_registrations_2022 <- party_registrations_2022 |>
  rename(TOTAL = X11)
```

```{r}
# Removals
removals_2022 <- read_csv("removals_2022.csv" , col_names = FALSE)

removals_2022 <- removals_2022 |>
rename(SUMMARY = X1)
removals_2022 <- removals_2022 |>
rename(DEM = X2)
removals_2022 <- removals_2022 |>
rename(REP = X3)
removals_2022 <- removals_2022 |>
rename(NLM = X4)
removals_2022 <- removals_2022 |>
rename(LIB = X5)
removals_2022 <- removals_2022 |>
rename(GRN = X6)
removals_2022 <- removals_2022 |>
rename(UNAF = X7)
removals_2022 <- removals_2022 |>
rename(OTH = X8)
removals_2022 <- removals_2022 |>
rename(TOTAL = X9)
removals_2022 <- removals_2022 |>
  filter(!is.na(DEM))

```

```{r}
# Changes
changes_2022 <- read_csv("changes_2022.csv" , col_names = FALSE)
changes_2022 <- changes_2022 |>
  rename(COUNTY = X1)
changes_2022 <- changes_2022 |>
  rename(ADDRESS = X2)
changes_2022 <- changes_2022 |>
  rename(NAME = X3)

```

```{r}
# Current registrations

current_2022 <- read_csv("current_2022.csv" , col_names = FALSE)
current_2022 <- current_2022 |>
  select(-X2, -X3, -X4, -X5, -X6, -X7, -X8, -X9, -X10, -X11, -X20,-X21)

current_2022 <- current_2022 |>
  rename(COUNTY = X1)
current_2022 <- current_2022 |>
  rename(DEM = X12)
current_2022 <- current_2022 |>
  rename(REP = X13)
current_2022 <- current_2022 |>
  rename(NLM = X14)
current_2022 <- current_2022 |>
  rename(LIB = X15)
current_2022 <- current_2022 |>
  rename(GRN = X16)
current_2022 <- current_2022 |>
  rename(UNAF = X17)
current_2022 <- current_2022 |>
  rename(OTH = X18)
current_2022 <- current_2022 |>
  rename(TOTAL = X19)

```

## Answer questions

Q1. Calculate the percentage of registered Democrats, Republicans and unaffiliated voters for the entire state in September 2024 and September 2022, and then calculate the percentage change for those three party categories (the formula for percentage change is easy to remember: (New-Old)/Old). How would you describe the changes in these partisan categories in a story? What's the most interesting thing about these changes that would be the lede of that story?

A1. I looked up online how to do percent change in R. 


The latest voter registration data for 2024 reveals changes in party affiliation compared to two years earlier. Democratic registrations have decreased slightly by 1.4%, dropping from 52.45% to 51.18% of the total voter base. Republican registrations have risen by 1.04%, increasing from 22.70% to 23.85%. Most strikingly, the percentage of unaffiliated voters has plummeted by a staggering 75.05%, dropping from 19.12% to just 4.96%.

Democrat drop, GOP gains, unaffiliated plunge, according to latest voter registration data. 



```{r}
total_current <- tibble(
  YEAR = c(2022, 2024),
  TOTAL = c(43999, 4231403),
  DEM = c(2250961, 2219417),
  REP = c(998905, 1009323 ),    
  UNAF = c(841573, 210000) 
)

total_current <- total_current %>%
  mutate(
    PERCENT_DEM = DEM / TOTAL * 100,
    PERCENT_REP = REP / TOTAL * 100,
    PERCENT_UNAF = UNAF / TOTAL * 100
  )

total_current <- total_current %>%
  mutate(
    CHANGE_DEM = (DEM - lag(DEM)) / lag(DEM) * 100,
    CHANGE_REP = (REP - lag(REP)) / lag(REP) * 100,
    CHANGE_UNAF = (UNAF - lag(UNAF)) / lag(UNAF) * 100
  )
total_current <- total_current %>%
filter(!is.na(CHANGE_DEM))
```

Q2. Among the Maryland jurisdictions with the most registered voters, which one had the biggest percentage increase in unaffiliated voters between 2022 and 2024? How did that county vote for governor in 2022 and for president in 2020? Contrast the partisan percentages of that county with the county that had the eighth-largest percentage increase in unaffiliated voters. You should display information from the data that shows the counties with the largest percentage increase first.

A2.Montgomery had the largest amount of total registered voters in 2024 and 2022. The biggest percentage increase in UNAF voters, however, was in PG County who has the second highest number of registered voters. They had around a 14% increase in UNAF voters from 2022 to 2024, which I just simply inputted into a % change calculator. 

Partisan data from PG, 2022: DEM:464963 REP:37836  , 2024: DEM: 456887 REP:39974
Partisan data from Frederick, 2022: DEM:77282 REP:68809 2024: DEM: 78789 REP: 68925

PG is simply just larger and has significantly more voters. Although they have more, Frederick has more of a population of Republicans. 

PG GOV. 2022 - Wes Moore
PG PRESIDENT 2024 - Joe Biden

```{r}
top_current <- current %>%
  group_by(COUNTY, UNAF) |>
  summarise(
  ) |>
  arrange(desc(UNAF))

top_current_2022 <- current_2022 %>%
  group_by(COUNTY, UNAF) %>%
  summarise() %>%
  arrange(desc(UNAF))

merged_data <- top_current %>%
  select(COUNTY, UNAF) %>%
  rename(UNAF_2024 = UNAF) %>%
  left_join(top_current_2022 %>% select(COUNTY, UNAF), by = "COUNTY") %>%
  rename(UNAF_2022 = UNAF)

merged_data <- merged_data %>%
  mutate(
    percentage_change = ((UNAF_2022 - UNAF_2024) / UNAF_2024) * 100
  )
merged_data|>
group_by(COUNTY, percentage_change)|>
  summarise()|>
  arrange(percentage_change)



```

Q3. Do the party percentages of new registrants (% Dem, % Rep, % Unaffiliated) from September 2024 and September 2022 match the overall percentage of partisan registration in the state at those times?

In other words, did new registrants mirror the overall partisan composition of the electorate, or were they different? Do any of the methods of registration vary significantly from the overall partisan pattern? Which method is worthy of further exploration?

A3.The party registration in PG county in 2022 of Dems more closely refers to the overall than REP. 
In 2024, this information is NA. 

I think it is more interesting to look at how the Republican does not closely refer to the overall because you could explore how a certain county votes. 


```{r}
party_percentage <- party_registrations %>%
  group_by(COUNTY, UNAF, REP, DEM) |>
  summarise(
  ) |>
  arrange()

party_percentage_2022 <- party_registrations_2022 %>%
  group_by(COUNTY, UNAF, REP, DEM) |>
  summarise(
  ) |>
  arrange()

party_registrations <- party_registrations %>%
  mutate(
    UNAF = as.numeric(as.character(UNAF)),
    REP = as.numeric(as.character(REP)),
    DEM = as.numeric(as.character(DEM))
  )


party_percentage <- party_registrations %>%
  group_by(COUNTY) %>%  
  summarise(
    TOTAL = sum(UNAF, REP, DEM), 
    PERCENT_UNAF = sum(UNAF) / TOTAL * 100, 
    PERCENT_DEM = sum(DEM) / TOTAL * 100,
    PERCENT_REP = sum(REP) / TOTAL * 100
  ) %>%
  arrange()

party_registrations_2022 <- party_registrations_2022 %>%
  mutate(
    UNAF = as.numeric(as.character(UNAF)),
    REP = as.numeric(as.character(REP)),
    DEM = as.numeric(as.character(DEM))
  )

party_percentage_2022 <- party_registrations_2022 %>%
  group_by(COUNTY) %>%  
  summarise(
    TOTAL = sum(UNAF, REP, DEM),
    PERCENT_UNAF = sum(UNAF) / TOTAL * 100,
    PERCENT_DEM = sum(DEM) / TOTAL * 100, 
    PERCENT_REP = sum(REP) / TOTAL * 100
  ) %>%
  arrange() 


```

Q4. Choose your own adventure! Using both the 2024 and 2022 data for one of the dataframes (changes, removals or new registrations), ask and answer a question that makes at least one comparison and requires code to answer. That could involve calculating a percentage, percentage change or the raw difference between two columns.

A4.

How many more Democratic voters were removed in September 2024 compared to September 2022?
1,848

```{r}

removals_2022 <- tibble(
  DEM = c(11533) )

removals <- tibble(
  DEM = c(9685) 
)

removals_2022_new <- removals_2022 %>% 
select(DEM)
removals_new <- removals %>% 
select(DEM)


removals_2022_new <- removals_2022_new %>% 
rename(DEM_2022 = DEM)
removals_new <- removals_new %>% 
rename(DEM_2024 = DEM)


dem_removal_change <- removals_new %>%
  bind_cols(removals_2022_new) %>% 
  mutate(DEM_CHANGE = DEM_2024 - DEM_2022) %>%  
  select(DEM_CHANGE) 

dem_removal_change




```

Q5. What's the best story idea or question you've seen as a result of the work you've done in this lab?

A5.

One interesting story idea that came from this lab is looking at how voter removals affect local elections. By comparing the rates of removals between 2022 and 2024 in different counties, we can see if there are patterns in who is being removed from the voter rolls. This could help us understand how these changes might impact voter turnout and representation in those areas, showing why it's important to protect people's voting rights.


```
