---
title: "lab_11"
author: "Derek Willis"
date: "2024-11-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries for working with data, including dates and column names, plus rvest.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
#| output: false
library(rvest)
library(tidyverse)
library(janitor)
library(ggplot2)
library(lubridate)
```


Let's get to scraping. We'll be working on collecting information about Maryland election results, and then we'll grab some congressional press releases. For Maryland races, we'll focus on Board of Education races using this CNS story as a guide: https://cnsmaryland.org/2024/11/08/md-conservatives-make-gains-in-school-board-races/. You should read it.

## Questions

**Q1**. Write code to scrape the table of unofficial results from Frederick County's Board of Education races (https://elections.maryland.gov/elections/2024/general_Results/gen_results_2024_by_county_11.html), producing a dataframe that contains the results of that race for each candidate and removing the total. You'll need to identify which table on the page contains the BOE results. All numbers should actually be numbers, including the percentage. Then make a bar chart of the results, noting that the top 3 candidates win.

**A1**

```{r}

url <- "https://elections.maryland.gov/elections/2024/general_Results/gen_results_2024_by_county_11.html"

tables <- url |> 
  read_html() |> 
  html_table()

boe_table <- tables[[9]]

boe_table <- boe_table |>
  clean_names() |>
  rename(
    early_voting = 2,
    election_day = 3,
    mail_in_ballot = 4,
    provisional = 5,
    total = 6,
    percentage = 7
  ) |>
  slice(-9) |> 
  mutate(
    early_voting = as.numeric(gsub(",", "", early_voting)),
    election_day = as.numeric(gsub(",", "", election_day)),
    mail_in_ballot = as.numeric(gsub(",", "", mail_in_ballot)),
    provisional = as.numeric(gsub(",", "", provisional)),
    total = as.numeric(gsub(",", "", total)),
    percentage = as.numeric(gsub("%", "", percentage))
  )

boe_table |> 
  ggplot(aes(x = reorder(name, percentage), y = percentage)) + 
  geom_bar(stat = "identity", fill = "pink") +  
  coord_flip() +  
  theme_minimal() +  
  labs(
    title = "Tight Race in Frederick County's Board of Education: Results",
    x = "Candidates",
    y = "Shared Vote %",
    subtitle = "Top 3 Candidates Win"
  )

```

**Q2** Next, let's scrape the list of press releases from Maryland's Office of the Public Defender, <https://www.opd.state.md.us/press-releases>. This isn't a table, so you'll need to use `html_elements()` and your browser's inspector and do some clean up on the results. The result should be a dataframe with three columns: title, url and date. HINT: you can extract the date from the title using lubridate OR you can use the `separate` function.

You should have 10 releases when finished, not 20.

Then, write code that finds the most recent release with the word "Campaign" in the title. What election does it refer to?

**A2** John King for Governor Campaign Cited for Authority Line Violations , Gov race

```{r}
press_url <- "https://osp.maryland.gov/category/press-releases"

html <- press_url |>
  read_html()
html

press_releases <- html |> #it is AHREF
  html_elements("a") 

urls <- tibble(
  title = press_releases %>% html_attr("title"), 
  url = press_releases %>% html_attr("href") #href here
) |> 
  mutate(
    date = mdy(title), #lubridate
    title = str_remove(title, "Permanent Link to .*?: ") #remove the extraneous info from the title, I had no clue how to do this so I looked it up on ChatGPT
) |> 
  drop_na() #remove NA so there are 10, not 20



campaign <- urls |> 
  filter(str_detect(title, "Campaign"))

campaign


```

**Q3** Sen. Ben Cardin, D-Maryland, has posted hundreds of press releases at <https://www.cardin.senate.gov/?post_type=press-releases>. It would be great to have all of them in a dataframe that has the following columns: date, title and url.

To do this, you will need to scrape the page's html and save that to a variable, and *then* extract the dates, titles and urls into *separate* dataframes using html_elements(). We turn a list into a dataframe using `as_tibble()`.

At the end, you'll have three dataframes that you want to combine into a single dataframe. When we want to combine the rows of identical dataframes, we used `bind_rows()`. If you were combining columns instead of rows, there's a similar function. Use it to put all of the dataframes together into a single one. You are combining columns, not rows.

When you're done, rename the columns so they make sense, then make sure the date column is an actual date.

Finally, tell me what questions you could ask of this data, and what other information about it would be useful to have. Be creative.

**A3** 
What is the most frequent word in the titles of press releases? This is newsworthy because identifying the most frequent word in press release titles can reveal the key topics or issues being prioritized by the politican. I would need the full text of the press release titles in my dataset. 

How many press releases mention a specific senator or public figure? This is newsworthy because it shows how often a senator or public figure is in the spotlight, revealing their role in current events.I’d need to identify specific senators or public figures in the titles or text of the press releases. If their names aren’t explicitly mentioned in my dataset, I might need to scrape the content of the press releases or have a list of names to search for in the text.

Which months have the most press releases? This is newsworthy because it shows when a senator or public figure is most active, which highlighs when politicans are most active. I’d need the date data for each press release to group them by month.

```{r}
url_cardin <- "https://www.cardin.senate.gov/?post_type=press-releases"

html_cardin <- url_cardin |>
  read_html()
url_cardin

cardin_dates <- html_cardin |> 
  html_elements("h5") |> 
  html_text() |> 
  mdy() |> 
  as_tibble() |> 
  rename(date = value)

cardin_titles <- html_cardin |> 
  html_elements("h3 a") |> 
  html_text() |> 
  as_tibble() |> 
  rename(title = value)

cardin_urls <- html_cardin |> 
  html_elements("h3 a") |> 
  html_attr("href") |> 
  as_tibble() |> 
  rename(url = value)

cardin_dates_titles_urls <- bind_cols(cardin_dates, cardin_titles, cardin_urls)

cardin_dates_titles_urls

```

